Counting evolutions: a simple foundation for thermodynamics

### 2021-01-12 Articles/2019-Entropy/Source/thermodynamics.tex
After review with Christine

\begin{abstract}
	In this work we present the core ideas at the heart of a new approach to the foundations of thermodynamics. The premise is to focus on the count of evolutions: the possible ways the state of a system can evolve in time. We see that in a deterministic process the evolutions can never diverge and in a reversible one they can never merge. This means that the count of evolutions per state can never decrease under a deterministic process and it is maximized at equilibrium. We show how this simple idea can be used as a straightforward foundation of thermodyanmics and how it relates to other notions of entropy in physics.

@@

This working draft summarizes the status of our work on a new approach to the foundations of thermodynamics and is intended to gather early feedback. Rather then derive thermodynamics bottom-up from a particular microscopic model, we want to derive it top-down from a set of requirements on the dynamics of the system. The results we find are, then, independent on the microscopic dynamics in the sense that they can in principle be realized in different ways. This is both more in line with the aims of the founders of thermodynamics and justifies why thermodynamic concepts have been successful in fields beyond physics. Moreover, in our approach thermodynamics is found as a specialization of a more general model, which we believe can be used as a common foundation for the different types mechanics.

In a nutshell, we characterize processes in general terms, as the description of all possible ways that the system can evolve in time, its possible evolutions, under a particular set of circumstances, as one can see in figures \ref{fig_single_evolution} through \ref{fig_with_equilibria}. Counting the possible evolutions plays a fundamental role, in much the same way that counting states plays a role in statistical mechanics. In a deterministic and reversible process, all evolutions that start in one state will end in the same state: the evolutions cannot merge or diverge. If the process is deterministic but not reversible, however, the same final state can be reached from different initial states, and the evolution will merge. If the deterministic process reaches an equilibrium, there will be no more merging. We define process entropy to be the logarithm of the count of evolutions per state, which we find to have the basic properties we associate with entropy: it is linear under system composition of independent systems, can never decrease under a deterministic process, increases if the process is non-reversible and is maximized at equilibrium. This gives us a conceptually crisp physically meaningful concept that is tied to irreversibility and equilibria in a very natural way.

There are many technical details that are required to make the above characterization formally rigorous, mostly due to the proper handling of infinities. However, those do not really change the overall intuition, which we believe is the main strength of the approach. Therefore, in this paper, we will only give a conceptual overview leaving the formal details to other papers. This will leave the key physical insight is more directly accessible, and also result in a narrative that can be used directly in an introductory course of thermodynamics.

@@

Not all evolutions will be possible under all conditions. Therefore we choose to characterize a \textbf{process} simply as the set of allowed evolutions, as we can see in figure \ref{fig_process}. Namely, a process tells us what can and cannot happen in a given set of circumstances. We also assume we have a measure $\mu$ that allows us to determine the size of a set of evolutions, to ``count'' evolutions.\footnote{The precise conceptual and mathematical definition for $\mu$ is source of a number of technical considerations which we (will) address in a separate work. While some issues, like unit definition, are important to the physics, most have to do with the proper handling of infinities or the conditions for existence and uniqueness of the measure. Overall, they would distract from the key insights.} A state $x_t$ at a time $t$ identifies a set of of compatible evolutions $A(x_t)$, those that pass through that state at that time. That is, $x(t) = x_t$. Using $\mu$ we can give a size to that set, which we can write as $\mu(A(x_t))$ or $\mu(x_t)$ for short.

@@

Note that the probability of transitioning from one state to another during a process is linked to the allowed evolutions and their count. In fact, calculating the probability of finding the state $x_{t+\Delta t}$ at $t+\Delta t$ given $x_t$ at $t$ means determining the fraction of evolutions that pass through $x_t$ that also pass through $x_{t+\Delta t}$. That is:
\begin{equation}
	P(x_{t+\Delta t} | x_t) = \frac{\mu(A(x_t) \cap A(x_{t+\Delta t}))}{\mu(A(x_t))}.
\end{equation}
Conceptually, the ability to count evolutions is equivalent to knowing all possible probability of transition.\footnote{Technically, a single probability measure is not enough to define all these transition at all scales, but the overall intuition remains.}

Having answered the question of how we characterize processes, we can categorize them. For example, we say a process is \textbf{deterministic} if the state at one time in enough to determine the state at all future times. We can see an example of such a process in figure \ref{fig_determinism}. In this case, if two different evolutions happen to reach the same state at a given time, they will also share all future states: evolutions can merge but cannot diverge. The count of evolutions per state over time can never decrease: $
\mu(x(t)) \leq \mu(x(t + \Delta t))$.

@@

\begin{figure}[h!]
	\includegraphics[width=\columnwidth]{images/Slide3.png}
	\caption{Deterministic process: the state one time determines the state at future times. Evolutions can merge but not diverge in time. The number of evolutions per state cannot decrease.}\label{fig_determinism}
\end{figure}

@@

%TODO: add a double evolution in some case.
\begin{figure}[h!]
	\includegraphics[width=\columnwidth]{images/Slide5.png}
	\caption{Determinism and reversibility. Evolutions cannot merge nor split. The number of evolutions per state cannot increase.} \label{fig_detrev}
\end{figure}

To finish, consider a \textbf{deterministic processes with equilibria}. We can see an example of such a process in figure \ref{fig_with_equilibria}. An equilibrium for a process is a state that, once it is reached, it is never left.\footnote{We will see later that this characterization of equilibrium is symplistic, but it will suffice for now.} In a process with equilibria all evolutions will reach, at some point, an equilibrium. In the out-of-equilibrium phase, evolutions merge until they reach equilibria and can no longer merge. This means that our measure $\mu$ is maximized at equilibria. If we care to characterize only the initial and final state, deterministic evolution simply means that the final equilibrium is a function of the initial state.

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{images/Slide6.png}
	\caption{Deterministic process with equilibria. Evolutions keep merging until equilibrium is reached. The number of evolutions per state increase and the maximum is reached at equilibria.}\label{fig_with_equilibria}
\end{figure}

We already reached an important result: with only a handful of simple definitions we have found a quantity, $\mu(x(t))$, that can never decrease in a deterministic process, stays the same if the process is reversible and is maximized at equilibrium. We have done so with no assumption on the type of system, the type of forces acting on the system, the internal composition of the system or the internal dynamics of the system. It knows nothing of uncertainty, disorder, information, statistical or probability distribution, phase space volumes. It only knows about evolutions and how to count them. The result is completely general and the justification is very straight forward. It is conceptually a perfect candidate for the thermodynamic entropy, though we have one questions left: how does it combine under system composition?

Suppose we now have two systems, with their respective state space $X$ and $Y$. The state space of the composite system is the set of all possible pairs $X \times Y$. A process defined on the composite system will come with a measure $\mu_{X \times Y}$ that enables us to count evolutions $\{x(t), y(t)\}$ of the composite system. In general, the evolution of one system will affect the evolution of the other system. This will not happen, however, if the systems are independent. That is, as shown in \ref{fig_independence}, for each possible evolution $x(t)$ of the first system and $y(t)$ of the second system, there will be a possible evolution $\{x(t), y(t)\}$ on the composite. In other words,
\begin{equation}
	\mu_{X \times Y} = \mu_X \mu_Y
\end{equation}
the measure factorizes. We say two system are \textbf{independent} at given time $t$ if the evolutions per state factorizes: $\mu_{X \times Y}(x(t), y(t)) = \mu_X (x(t)) \mu_Y (y(t))$.

@@

We define the \textbf{process entropy} as $S = \log(\mu)$. Since the logarithm is a monotonic function, the process entropy will preserve monotonicity and maximization of $\mu$ under deterministic process with equilibria. Which means:
\begin{equation}\label{ov_entropy_increases}
S(x(t)) \leq S(x(t + \Delta t))
\end{equation}
In addition, the process entropy will be additive for independent systems.\footnote{As before there are a number of technical details that need to be addressed, most of all the units for $\mu$ which will influence the 0 point for $S$. We (will) address in the full framework.}

@@

If the system is a composite there are additional considerations. Different processes may put different degree of correlation between subsystems, therefore the same equilibria for the subsystems may correspond to different processes entropy values for the composite. But if the subsystems are independent, the composite will correspond to a unique entropy value: the product.

We can then assign a unique process entropy to states of equilibria and to composite state of independent equilibria. In these conditions we define the \textbf{state entropy} $S(x)$ which corresponds to the process entropy that the system reaches at equilibrium. But why would we want to assign a unique process entropy to states in the first place? Why would this be interesting? There are three main reasons.

The first is process combination. Suppose that we are studying a system and we change the circumstances surrounding it. It first evolves according to $A$ and we switch to process $B$. If $x(t)$ is an evolution of the combined process $AB$, then for $t\leq t_0$ it must be an evolution for $A$ and for $t\geq t_0$ it must be an evolution for $B$. Moreover, the number of total evolutions must remain the same. Therefore the state $x(t_0)$, the state where we are joining the evolutions, must have a well defined entropy that is independent on the process, or not all the evolutions would match. If we want to mix and match from a set of possible processes, then the process entropy at the junction must be only a function of the state.

The second reason is that, upon close inspection, measurements are never defined at a instant of time. Therefore $x(t)$, the state at an instant of time, is really a shorthand for $x[(t - dt/2, t + dt/2)]$, the state defined over a small open interval $dt$ around $t$. We can talk about a single state within that interval only if all state variables do not change and the system is fully characterized by those state variables alone. Therefore writing $x(t)$ implicitly assumes two things: that the system is an equilibrium of the processes that happen at a faster scale of our measurements of said system; that further description (e.g. the internal dynamics of the constituents at a finer scale) is not needed to describe the process.

@@

All three insights points toward the same directly: the very notion of state seems to be intrinsically associated to the notion of equilibrium and state entropy. This would suggest that all states for all systems are, in a sense, equilibria. But what is exactly this sense?  What does it mean that a mechanical system, classical or quantum, is at equilibrium?

Even in thermodynamics, equilibrium is usually an ill-defined concepts. It is typically associated the one coming from dynamical systems (i.e. the system no longer changes), but this doesn't quite work. In fact, it is not even relativistically consistent. If we have a volume of gas in thermodynamic equilibrium, we would like to say it is in equilibrium for any observer, not just the one in its rest frame. A booster observer, then, would see an equilibria where the position is indeed changing. In presence of gravitational forces, this motion would not even be linear. The notion of equilibrium as ``nothing is changing'', then, is not quite right and should be amended. But how? The previous discussion would hint toward system independence, but why system independence should be an equilibrium?

The insight is that we can define something as a system only when we are able to take it apart from the rest and study it independently. Therefore we need, at least conceptually, a process that takes that something and makes it independent from the rest, decouples it, makes it autonomous. Our ability to define a system implicitly assumes the existence of such ``decoupling'' process. Note that if the system is already independent, the decoupling process does nothing: in this sense independent systems are an equilibria of a decoupling process.

This ``decoupling'' process is what maximizes entropy. If the system is truly rendered independent, then we cannot know more than the description given by the state at that particular time. We cannot know how the system came to be in that state, we cannot know anything more about the subsystems, other systems or the environment than what the state of the system strictly tell us. All possible ways that the state can be realized must be realized in a possible evolution, or we could infer something more.  When the system is independent, the states tell us the least possible information about everything else and the evolution of anything else.

The definition of a particular system, then, is conditional on the existence of particular processes: those that render the system well defined. We can talk about the motion of a ball because we are on earth at 20 Celsius. If we were on the surface of the sun, we would not be able to talk about the ball in the first place. To define a system requires us to define a boundary, which should remain the same during the process. In mathematical terms, the state space must be a symmetry (i.e. an invariant, an equilibrium) of a group (i.e. a set of transformation, a set of processes).

The full mathematical development of these concepts is left for another work, but the conceptual foundations are simple and direct. Grouping quantities into cohesive units that ``belong together'', into systems, comes with a set of practical requirements: we need a way to prepare and study them independently. This notion of ``independence equilibrium'' is what we will use as a foundation for ``thermodynamic equilibrium'', and is generalized to all systems. It is the necessary and sufficient condition to assign state entropies. In this light, entropy maximization is not a universal law of the universe, but a necessary requirement to build models. We can only study the state of objects that allow process entropy maximization. What remains is to answer one last fundamental question: is state entropy at all related to thermodynamic entropy?

@@

We saw how the notion of state as independence equilibrium came naturally when considering process composition of a single system. When the process changes, we need to reconcile the count of evolution at the junction. We now want to consider the scenario of process composition of composite systems. The idea is that we want to describe the situation in which two systems come together, evolve as a composite for some time, and then proceed to evolve separately. In this case, at each juncture we have to reconcile the evolutions of both the whole and each of the part. We need to have an independence equilibrium at both levels. What we'll find is that thermodynamics studies this case. But let's proceed in order.

We say a system is in \textbf{thermodynamic equilibrium} if:
\begin{itemize}
	\item it is at equilibrium
	\item all its parts are in thermodynamic equilibrium
	\item if another system joined with a part of the system in thermodynamic equilibrium is in thermodynamic equilibrium, then the whole is in thermodynamic equilibrium.
\end{itemize}
We say two or more systems are in thermodynamic equilibrium if their composite is in thermodynamic equilibrium. A \textbf{thermodynamic process} is a process with equilibria where the final state is a thermodynamic equilibrium. We stress that we are not saying that two systems, in general, always allow for thermodynamic equilibria. However, we contend that these are the type of processes we can mix and match to create composite processes of composite systems. What we want to explore is what happens if we had such systems and process, at least to a first approximation.

To get a feel for the definition, let us rederive the fact that thermodynamic equilibrium is transitive, also known as zeroth law. Suppose system $A$ is in thermodynamic equilibrium with system $B$ and system $B$ is in thermodynamic equilibrium with system $C$. Then the composite $B \times C$ is in thermodynamic equilibrium by definition. Since $A$ is in thermodynamic equilibrium with $B$, a part of $B \times C$, it is also in thermodynamic equilibrium with $B \times C$. Therefore $A \times B \times C$ is in thermodynamic equilibrium. But $A \times C$ is a part of $A \times B \times C$, so it is in thermodynamic equilibrium. Which means $A$ is in thermodynamic equilibrium with $C$.

The characterization of thermodynamic equilibrium we give is manifestly a part-whole relationship. It extends the notion of independence equilibrium to a composite system so that the equilibrium of the whole and the equilibrium of its parts are essentially the same equilibrium. We find that clarifying this kind of part-whole relationships make for a good conceptual starting point not only for thermodynamics, but also for the other branches of physics. This should not come as a surprise as reductionism (i.e. explaining the whole through its parts) plays an important role in science.

Another important question for a part-whole relationships is: how are the state variables of the whole related to the state variables of the parts? We define a \textbf{thermodynamic system} one that follows these rules: 
\begin{itemize}
	\item states (i.e. independence equilibria) are identified by a set of extensive quantities (i.e. additive under system composition)\footnote{It may be possible that the existence of extensive quantities can be derived from the composability of thermodynamic equilibria. Note, in fact, that we typically construct measurement scales of quantities (such as energy, mass, volume, distance, ...) by preparing unit samples and then adding or dividing them. Assuming the existence of thermodynamic equilibria may enable us similar constructions without extra assumptions. (XXX: any work related to this?) Nevertheless, for this work, we consider the existence of extensive quantities an extra assumption.
	}
	\item one such quantity is called internal energy
	\item internal energy will be conserved under any deterministic process for that system.
\end{itemize}
We will denote $U$ as the internal energy and $x^i$ as the remaining extensive quantities.

For a thermodynamic system, then, the state entropy can be expressed as $S(U, x^i)$ a function of the extensive quantities. This is simply the equation of state in entropic form, which is the starting point of Gibbs formulation of thermodynamics. [CITE] We can therefore define the relationships
\begin{align}
	\beta &= \frac{\partial S}{\partial U} = \frac{1}{k_B T} \;\;\;\;\;\; - \beta X_i = \frac{\partial S}{\partial x^i}
\end{align}
where $k_B$ is Boltzmann's constant. We find:
\begin{align}
dS &= \frac{\partial S}{\partial U} dU + \frac{\partial S}{\partial x^i} dx^i \\
&= \beta dU - \beta X_i dx^i \\
dU &= k_B T dS + X_i dx^i
\end{align}
which are the usual thermodynamic relationships, except that, for us, the state entropy $S$ dimensionless..

@@

A natural question, based on the previous definition, is: to what extend thermodynamic processes can change the energy and the entropy of each system? Intuitively, there are two ``orthogonal'' directions for a change of energy: constant entropy and maximal entropy change. To study these components, it is useful to consider the following special thermodynamic systems.

@@

We say $M$ is a \textbf{mechanical system} if the state entropy is a constant for all states. For a thermodynamic system that is also mechanical, we therefore have  $S(U, x^i) = k_M$ for all its states. A mechanical system cannot change entropy and all states can be connected through a deterministic and reversible process. We have:
\begin{equation}
\begin{aligned}
dU &= k_B T dS + X_i dx^i = X_i dx^i.
\end{aligned}
\end{equation}
The energy of the mechanical system is effectively \emph{stored} in one of the other state variables and can be later retrieved. We call \textbf{work} the energy exchanged by a mechanical system and we note $W=\Delta U$ as the energy gained by the system during a thermodynamic process.\footnote{Note that $\frac{\partial S_M}{\partial U} = 0 = \beta_M = \frac{1}{k_B T_M}$ which means the thermodynamic ``temperature'' of a mechanical system is infinite. While this ``temperature'' has little to do with thermometers, the fact that we can always transform all work into heat but not vice-versa is just a specific case of being able to move energy from higher to lower temperature.}

We can study how changes of energy and entropy are related to each other. Consider a system composed of a generic system $A$, a thermal reservoir $R$ and a mechanical system $M$. In this setting, work represents the energy exchanged at constant entropy while heat represents energy exchanged associated with entropy change.

>>>>

\begin{abstract}
	In this work we present the core ideas at the heart of a new approach to the foundations of thermodynamics. The premise is to focus on the count of evolutions: the possible ways the state of a system can evolve in time. We see that in a deterministic process the evolutions can never diverge and in a reversible one they can never merge. This means that the count of evolutions per state can never decrease under a deterministic process and it is maximized at equilibrium. We show how this simple idea can be used as a straightforward foundation for thermodynamics and how it relates to other notions of entropy in physics.

@@

This working draft summarizes the status of our work on a new approach to the foundations of thermodynamics and is intended to gather early feedback. Rather then derive thermodynamics bottom-up from a particular microscopic model, we want to derive it top-down from a set of requirements on the dynamics of the system. The results we find are, then, independent of the microscopic dynamics in the sense that they can in principle be realized in different ways. This is both more in line with the aims of the founders of thermodynamics and justifies why thermodynamic concepts have been successful in fields beyond physics. Moreover, in our approach thermodynamics is found as a specialization of a more general model, which we believe can be used as a common foundation for the different types of mechanics.

In a nutshell, we characterize processes in general terms, as the description of all possible ways that the system can evolve in time, its possible evolutions, under a particular set of circumstances, as one can see in figures \ref{fig_single_evolution} through \ref{fig_with_equilibria}. Counting the possible evolutions plays a fundamental role, in much the same way that counting states plays a role in statistical mechanics. In a deterministic and reversible process, there is one to one correspondence between initial and final states: the evolutions cannot merge or diverge. If the process is deterministic but not reversible, however, the same final state will be reached from different initial states, and the evolutions will merge. If the deterministic process reaches an equilibrium, there will be no more merging. We define process entropy to be the logarithm of the count of evolutions per state, which we find to have the basic properties we associate with entropy: it is linear under system composition of independent systems, can never decrease under a deterministic process, increases if the process is non-reversible and is maximized at equilibrium. This gives us a conceptually crisp physically meaningful concept that is tied to irreversibility and equilibria in a very natural way.

There are many technical details that are required to make the above characterization formally rigorous, mostly due to the proper handling of infinities. However, those do not really change the overall intuition, which we believe is the main strength of the approach. Therefore, in this paper, we will only give a conceptual overview leaving the formal details to other papers. This will leave the key physical insight more directly accessible, and also result in a narrative that can be used directly in an introductory course of thermodynamics.

@@

Not all evolutions will be possible under all conditions. Therefore we choose to characterize a \textbf{process} simply as the set of allowed evolutions, as we can see in figure \ref{fig_process}. Namely, a process tells us what can and cannot happen in a given set of circumstances. We also assume we have a measure $\mu$ that allows us to determine the size of a set of evolutions, to ``count'' evolutions.\footnote{The precise conceptual and mathematical definition for $\mu$ is a source of a number of technical considerations which we (will) address in a separate work. While some issues, like unit definition, are important to the physics, most have to do with the proper handling of infinities or the conditions for existence and uniqueness of the measure. Overall, they would distract from the key insights.} A state $x_t$ at a time $t$ identifies a set of of compatible evolutions $A(x_t)$, those that pass through that state at that time. That is, $x(t) = x_t$. Using $\mu$ we can give a size to that set, which we can write as $\mu(A(x_t))$ or $\mu(x_t)$ for short.

@@

Note that the probability of transitioning from one state to another during a process is linked to the allowed evolutions and their count. In fact, calculating the probability of finding the state $x_{t+\Delta t}$ at $t+\Delta t$ given $x_t$ at $t$ means determining the fraction of evolutions that pass through $x_t$ that also pass through $x_{t+\Delta t}$. That is:
\begin{equation}
	P(x_{t+\Delta t} | x_t) = \frac{\mu(A(x_t) \cap A(x_{t+\Delta t}))}{\mu(A(x_t))}.
\end{equation}
Conceptually, the ability to count evolutions is equivalent to knowing all possible transition probabilities.\footnote{Technically, a single probability measure is not enough to define all these transitions at all scales, but the overall intuition remains.}

Having answered the question of how we characterize processes, we can categorize them. For example, we say a process is \textbf{deterministic} if the state at one time is enough to determine the state at all future times. We can see an example of such a process in figure \ref{fig_determinism}. In this case, if two different evolutions happen to reach the same state at a given time, they will also share all future states: evolutions can merge but cannot diverge. The count of evolutions per state over time can never decrease: $
\mu(x(t)) \leq \mu(x(t + \Delta t))$.

@@

\begin{figure}[h!]
	\includegraphics[width=\columnwidth]{images/Slide3.png}
	\caption{Deterministic process: the state at one time determines the state at future times. Evolutions can merge but not diverge in time. The number of evolutions per state cannot decrease.}\label{fig_determinism}
\end{figure}

@@

\begin{figure}[h!]
	\includegraphics[width=\columnwidth]{images/Slide5.png}
	\caption{Determinism and reversibility. Evolutions cannot merge nor split. The number of evolutions per state cannot increase.} \label{fig_detrev}
\end{figure}

To finish, consider a \textbf{deterministic processes with equilibria}, like the one in figure \ref{fig_with_equilibria}. An equilibrium for a process is a state that, once it is reached, it is never left.\footnote{We will see later that this characterization of equilibrium is simplistic, but it will suffice for now.} In a process with equilibria all evolutions will reach, at some point, an equilibrium. In the out-of-equilibrium phase, evolutions merge until they reach equilibria and can no longer merge. This means that our measure $\mu$ is maximized at equilibria. If we care to characterize only the initial and final state, deterministic evolution simply means that the final equilibrium is a function of the initial state.

\begin{figure}[h]
	\includegraphics[width=\columnwidth]{images/Slide6.png}
	\caption{Deterministic process with equilibria. Evolutions keep merging until equilibrium is reached. The number of evolutions per state increases and the maximum is reached at equilibria.}\label{fig_with_equilibria}
\end{figure}

We already reached an important result: with only a handful of simple definitions we have found a quantity, $\mu(x(t))$, that can never decrease in a deterministic process, stays the same if the process is also reversible and is maximized at equilibrium. We have done so with no assumption on the type of system, the type of forces acting on the system, the internal composition of the system or the internal dynamics of the system. It knows nothing of uncertainty, disorder, information, statistical or probability distribution, phase space volumes. It only knows about evolutions and how to count them. The result is completely general and the justification is very straight forward. It is conceptually a perfect candidate for the thermodynamic entropy, though we have one questions left: how does it combine under system composition?

Suppose we now have two systems, with their respective state spaces $X$ and $Y$. The state space of the composite system is the set of all possible pairs $X \times Y$. A process defined on the composite system will come with a measure $\mu_{X \times Y}$ that enables us to count evolutions $\{x(t), y(t)\}$ of the composite system. In general, the evolution of one system will affect the evolution of the other system. This will not happen, however, if the systems are independent. That is, as shown in \ref{fig_independence}, for each possible evolution $x(t)$ of the first system and $y(t)$ of the second system, there will be a possible evolution $\{x(t), y(t)\}$ on the composite. In other words,
\begin{equation}
	\mu_{X \times Y} = \mu_X \mu_Y
\end{equation}
the measure factorizes. We say two systems are \textbf{independent} at a given time $t$ if the evolutions per state factorizes: $\mu_{X \times Y}(x(t), y(t)) = \mu_X (x(t)) \mu_Y (y(t))$.

@@

We define the \textbf{process entropy} as $S = \log(\mu)$. Since the logarithm is a monotonic function, the process entropy will preserve monotonicity and maximization of $\mu$ under a deterministic process with equilibria. This means:
\begin{equation}\label{ov_entropy_increases}
S(x(t)) \leq S(x(t + \Delta t))
\end{equation}
In addition, the process entropy will be additive for independent systems.\footnote{As before there are a number of technical details that need to be addressed, most of all the units for $\mu$ which will influence the 0 point for $S$. We will address this in the full framework.}

@@

If the system is a composite there are additional considerations. Different processes may put different degrees of correlation between subsystems, therefore the same equilibria for the subsystems may correspond to different process entropy values for the composite. But if the subsystems are independent, the composite will correspond to a unique entropy value: the product.

We can then assign a unique process entropy to states of equilibria and to composite states of independent equilibria. Under these conditions we define the \textbf{state entropy} $S(x)$ which corresponds to the process entropy that the system reaches at equilibrium. But why would we want to assign a unique process entropy to states in the first place? Why would this be interesting? There are three main reasons.

The first is process combination. Suppose that we are studying a system and we change the circumstances surrounding it. It first evolves according to $A$ and at $t_0$ we switch to process $B$. If $x(t)$ is an evolution of the combined process $AB$, then for $t\leq t_0$ it must be an evolution for $A$ and for $t\geq t_0$ it must be an evolution for $B$. Moreover, the number of total evolutions must remain the same. Therefore the state $x(t_0)$, the state where we are joining the evolutions, must have a well defined entropy that is independent of the process, or not all the evolutions would match. If we want to mix and match from a set of possible processes, then the process entropy at the junction must be only a function of the state.

The second reason is that, upon close inspection, measurements are never defined at an instant of time. Therefore $x(t)$, the state at an instant of time, is really a shorthand for $x[(t - dt/2, t + dt/2)]$, the state defined over a small open interval $dt$ around $t$. We can talk about a single state within that interval only if no state variables change and the system is fully characterized by those state variables alone. Therefore writing $x(t)$ implicitly assumes two things: that the system is an equilibrium of the processes that happen at a faster scale than our measurements of said system; that further description (e.g. the internal dynamics of the constituents at a finer resolution) is not needed to describe the process.

@@

All three insights point toward the same direction: the very notion of state seems to be intrinsically associated to the notion of equilibrium and state entropy. This would suggest that all states for all systems are, in a sense, equilibria. But what exactly is this sense?  What does it mean that a mechanical system, classical or quantum, is at equilibrium?

Even in thermodynamics, equilibrium is usually an ill-defined concept. It is typically associated with the one coming from dynamical systems (i.e. the system no longer changes), but this doesn't quite work. In fact, it is not even relativistically consistent. If we have a volume of gas in thermodynamic equilibrium, we would like to say it is in equilibrium for any observer, not just the one in its rest frame. A boosted observer, then, would see an equilibrium where the position is indeed changing. In the presence of gravitational forces, this motion would not even be linear. The notion of equilibrium as ``nothing is changing'', then, is not quite right and should be amended. But how? The previous discussion would hint toward system independence, but why should system independence imply an equilibrium?

The insight is that we can define something as a system only when we are able to take it apart from the rest and study it independently. Therefore we need, at least conceptually, a process that takes that something and makes it independent from the rest, decouples it, makes it autonomous. Our ability to define a system implicitly assumes the existence of such a ``decoupling'' process. Note that if the system is already independent, the decoupling process does nothing: in this sense independent systems are an equilibrium of a decoupling process.

This ``decoupling'' process is what maximizes entropy. If the system is truly rendered independent, then we cannot know more than the description given by the state at that particular time. We cannot know how the system came to be in that state, we cannot know anything more about the subsystems, other systems or the environment than what the state of the system strictly tell us.   When the system is independent, the states tell us the least possible information about everything else and the evolution of anything else.

The definition of a particular system, then, is conditional on the existence of particular processes: those that render the system well defined. We can talk about the motion of a ball because we are on earth at 20 Celsius. If we were on the surface of the sun, we would not be able to talk about the ball in the first place. To define a system requires us to define a boundary, which should remain the same during the process. In mathematical terms, the state space must be a symmetry (i.e. an invariant, an equilibrium) of a group (i.e. a set of transformations, a set of processes).

The full mathematical development of these concepts is left for another work, but the conceptual foundations are simple and direct. Grouping quantities into cohesive units that ``belong together'', into systems, comes with a set of practical requirements: we need a way to prepare and study them independently. This notion of ``independence equilibrium'' is what we will use as a foundation for ``thermodynamic equilibrium'', and is generalized to all systems. It is the necessary and sufficient condition to assign state entropies. In this light, entropy maximization is not a universal law of the universe, but a necessary requirement to build models. We can only study the state of objects that allow process entropy maximization. What remains is to answer one last fundamental question: is this notion of state entropy at all related to thermodynamic entropy?

@@

We saw how the notion of state as independence equilibrium came naturally when considering process composition of a single system. When the process changes, we need to reconcile the count of evolution at the junction. We now want to consider the scenario of process composition of composite systems. The idea is that we want to describe the situation in which two systems come together, evolve as a composite for some time, and then proceed to evolve separately. In this case, at each juncture we have to reconcile the evolutions of both the whole and each of the parts. We need to have an independence equilibrium at both levels. What we'll find is that thermodynamics studies this case. But let's proceed in order.

If our goal is to characterize part-whole relationships, a crucial question is: how are the state variables of the whole related to the state variables of the parts? We define a \textbf{thermodynamic system} one that follows these rules: 
\begin{itemize}
	\item states (i.e. independence equilibria) are identified by a set of extensive quantities (i.e. additive under system composition)\footnote{It may be possible that the existence of extensive quantities can be derived from the composability of thermodynamic equilibria. Note, in fact, that we typically construct measurement scales of quantities (such as energy, mass, volume, distance, ...) by preparing unit samples and then adding or dividing them. Assuming the existence of thermodynamic equilibria may enable us similar constructions without extra assumptions. (XXX: any work related to this?) Nevertheless, for this work, we consider the existence of extensive quantities an extra assumption.
	}
	\item one such quantity is called internal energy
	\item internal energy will be conserved under any deterministic process for that system.
\end{itemize}
We will denote $U$ as the internal energy and $x^i$ as the remaining extensive quantities.

For a thermodynamic system, then, the state entropy can be expressed as $S(U, x^i)$ a function of the extensive quantities. This is simply the equation of state in entropic form, which is the starting point of Gibbs formulation of thermodynamics. [CITE] We can therefore define the relationships
\begin{align}
\beta &= \frac{\partial S}{\partial U} = \frac{1}{k_B T} \;\;\;\;\;\; - \beta X_i = \frac{\partial S}{\partial x^i}
\end{align}
where $k_B$ is Boltzmann's constant. We find:
\begin{align}
dS &= \frac{\partial S}{\partial U} dU + \frac{\partial S}{\partial x^i} dx^i \\
&= \beta dU - \beta X_i dx^i \\
dU &= k_B T dS + X_i dx^i
\end{align}
which are the usual thermodynamic relationships, except that, for us, the state entropy $S$ is dimensionless.

As we need to characterize states of both whole and parts, we need to characterize the part-whole relationship of the equilibrium. We say that a thermodynamic system $A$ in \textbf{thermodynamic equilibrium} has the following properties:
\begin{itemize}
	\item the system $A$ is in an independence equilbrium
	\item all parts of $A$ are in thermodynamic equilibrium
	\item if another composite system $B$ is in thermodynamic equilibrium and $A$ and $B$ share a part, then the composite of $A$ and $B$ is in thermodynamic equilibrium.
\end{itemize}
We say two or more systems are in thermodynamic equilibrium with each other if their composite is in thermodynamic equilibrium.\footnote{To understand the mathematical structure a bit better, the set of all systems $S$ is a Boolean algebra. Element $0$ represents ``nothing'' while element $1$ represents ``everything''. The join $\OR$ operation is system composition while the meet $\AND$ returns the smallest system contained by the two. Therefore $A \OR 1 = 1$: any system composed with everything gives us everything. As a lattice, the partial order $\leq$ is system inclusion. A thermodynamic equilibrium essentially identifies a topmost system, with all the subsystem and only the subsystems as part of the same equilibrium. That is, for every thermodynamic equilibrium there exists a system $A$ such that every $B \leq A$ is in the same equilibrium while every $C \nleq A$ is not in the same equilibrium.} A \textbf{thermodynamic process} is a process with equilibria where the final state is a thermodynamic equilibrium. We stress that we are not saying that two systems, in general, always allow for thermodynamic equilibria. However, we contend that these are the type of processes we can mix and match to create composite processes of composite systems. What we want to explore is what happens if we had such systems and processes, at least to a first approximation.

To get a feel for the definition, let us rederive the fact that thermodynamic equilibrium is transitive, also known as the zeroth law. Suppose system $A$ is in thermodynamic equilibrium with system $B$ and system $B$ is in thermodynamic equilibrium with system $C$. Then the composites $A \times B$ and $B \times C$ are respectively in thermodynamic equilibrium. Since they share a part, $B$, this means that $A \times B \times C$ is in thermodynamic equilibrium. But $A \times C$ is a part of $A \times B \times C$, so it is in thermodynamic equilibrium. Which means $A$ is in thermodynamic equilibrium with $C$.

The characterization of thermodynamic equilibrium we give is manifestly a part-whole relationship. It extends the notion of independence equilibrium to a composite system so that the equilibrium of the whole and the equilibrium of its parts are essentially the same equilibrium. We find that clarifying this kind of part-whole relationships makes for a good conceptual starting point not only for thermodynamics, but also for the other branches of physics. This should not come as a surprise as reductionism (i.e. explaining the whole through its parts) plays an important role in science.

@@

A natural question, based on the previous definitions, is: to what extent can thermodynamic processes change the energy and the entropy of each system? Intuitively, there are two ``orthogonal'' directions for a change of energy: constant entropy and maximal entropy change. To study these components, it is useful to consider the following special thermodynamic systems.

@@

We say $M$ is a \textbf{mechanical system} if the state entropy is a constant for all states. For a thermodynamic system that is also mechanical, we therefore have  $S(U, x^i) = k_M$ for all its states. A mechanical system cannot change entropy and all states can be connected through a deterministic and reversible process. We have:
\begin{equation}
\begin{aligned}
dU &= k_B T dS + X_i dx^i = X_i dx^i.
\end{aligned}
\end{equation}
The energy of the mechanical system is effectively \emph{stored} in one of the other state variables and can be later retrieved. We call \textbf{work} the energy exchanged by a mechanical system and we note $W=\Delta U$ as the energy gained by the system during a thermodynamic process.\footnote{Note that $\frac{\partial S}{\partial U} = 0 = \beta = \frac{1}{k_B T}$ which means the thermodynamic ``temperature'' of a mechanical system is infinite. While this ``temperature'' has little to do with thermometers, the fact that we can always transform all work into heat but not vice-versa is just a specific case of being able to move energy from higher to lower temperature.}

We can study how changes of energy and entropy are related to each other. Consider a system composed of a generic system $A$, a thermal reservoir $R$ and a mechanical system $M$. In this setting, work represents the energy exchanged at constant entropy while heat represents exchanged energy associated with entropy change.
